White Dwarf Mergers on Adaptive Meshes\\ II. Nuclear Reactions

### 2017-10-07 papers/wdmerger_II/paper.tex
Simplify discussion of burning modes

We define several burning modes that determine how $T$ and $e$ are evolved
during a nuclear burn. In a hydrostatic burn, which we call burning mode 0,
we keep $\rho$ and $T$ fixed throughout, and use 
the energy released at the end to compute a final temperature that is
thermodynamically consistent with the new internal energy. By contrast,
in a self-heating burn (mode 1), we allow the temperature to evolve in response
to the burning (see\footnote{In the cited paper, a term based on the
thermodynamic chemical potential was included; we now believe
that it is incorrect to include such a term in the burn, since it
automatically sums to zero analytically.} \citet{maestro3}):
\begin{equation}
  \frac{dT}{dt} = \frac{1}{c_V}\frac{\partial e}{\partial t}
\end{equation}
(Although $T$ evolves during the burn so that the integration is physically
accurate, as in the hydrostatic method we discard the final value
for $T$ at the end of the burn and recompute a temperature for the zone that is
consistent with its new internal energy.) Here $c_V$ is the specific heat at
constant volume, which is provided by the equation of state.  During this burn,
we can keep $c_V$ constant using its initial value, or at each step we
can choose to re-evaluate the equation of state using the latest value of $(\rho, T)$.
The latter is more expensive but also more accurate, and we use it in this paper.
In practice we find that the cost is small in comparison to the more expensive
parts of the calculation, and it can significantly speed up convergence near NSE.
A third option (mode 2) presented by \citet{raskin:2010} is a so-called ``hybrid'' mode.
In this mode, by default we do a hydrostatic burn. If that burn fails, or if the net
energy change is negative, we do the burn again in self-heating mode. A final option (mode 3)
is a burn that limits the changes due to a burn to avoid numerically unstable burning.
This mode is discussed in \autoref{sec:unstable_burning} and we will call it a
``suppressed'' burn for the remainder of this work. All four options
are implemented in our burner software. The simulations shown in this work all
use the self-heating mode unless otherwise specified.

@@

Our primary metric for this test is the amount of $^{56}$Ni generated in the collision,
as this is the parameter most directly related to the observable quantities of interest
for Type Ia supernovae. We collect information at the end of every timestep about the
total amount of nickel on the grid (in solar masses), and we use the maximum value of
this nickel mass. Plotfiles of a few relevant quantities (density, temperature, velocity,
pressure, sound speed, and the ratio $\tau_s / \tau_e$) are created every 0.01 seconds.
The timesteps are limited to ensure that every multiple of this plotfile interval is
reached exactly, which implies an effective maximum timestep of 0.01 seconds.

@@

In \autoref{sec:burner} we observed that there are several alternatives to the traditional
self-heating approach in a burn. These alternatives are relevant to a collision problem because,
as pointed out by \cite{raskin:2010} and \cite{kushnir:2013}, it is possible for a typical
zone in our simulation to release a very large amount of energy in a burn before it is cooled
off by a subsequent hydrodynamic expansion step, leading to the numerically unstable burning
problem described in \autoref{sec:unstable_burning}. \autoref{table:burningmode} lists nickel
production for the four burning modes described earlier. The hydrostatic burn does produces
slightly more nickel than the self-heating burn, consistent with the idea that by limiting the
energy release from early carbon burning, we can forestall a detonation which is numerically
spurious. Yet the difference is only about 5\%. The reason is that the hydrostatic burn controls
the energy release during the burn, but \textit{not} what happens in the rest of the hydrodynamics
step. And in the hydrodynamics step, the \castro\ algorithm performs multiple EOS calls to ensure
that the temperature is synchronized with the internal energy at various points in the update.
So in practice there is a temperature change due to the burn for our problem. This seems inevitable
for any hydro code unless the hydrodynamics update includes an explicit equation for $T$ that is
independent of what is happening for the internal energy. Similarly, the hybrid burn does not
affect the total nickel production relative to the hydrostatic burn. The hybrid burn only changes
the behavior for zones that have a net negative energy release from the hydrostatic burn, which
occurs only after we have burned to NSE.

\input{plots/burning_mode_m_P_0.64_m_S_0.64.tbl}

In contrast, the suppressed burning mode (modeled after \citet{kushnir:2013}) produces only about
$2 \times 10^{-4}\ \msolar$ of $^{56}$Ni. In a suppressed burn, we ensure that at every evaluation
of the right-hand-side vector for the nuclear network integration, all quantities are scaled by the
same factor, with this factor ensuring that the energy release is not large enough to permit a
spurious detonation. This seems to be equivalent to the method used in \citet{kushnir:2013}. We
reproduce their claim that this suppresses the detonation until later in the collision, after more
material has approached the stalled shock. We also observe that, for a given resolution, the simultaneous
detonations occur slightly outside the initial contact point, near the edges of the stalled shock. They
establish as the main source of error in simulations that produce too little nickel. However, we find
that the resulting detonations are not strong enough to convert a significant amount of material to NSE
conditions; instead, only QSE conditions are reached near the center of the collision, leaving a significant
amount of silicon and sulfur on the domain that is not further processed. In order to ensure that the
very different results we obtained were not a result of the specific choice we made for how to suppress
the burn, we tried varying the approach discussed in \autoref{sec:burner}. First, we limited based on
the current value of the internal energy and sound speed rather than the fixed, initial value during a burn
(this should be a less severe limiter). This did not yield a significant difference in the outcome. Second,
we also tried applying the limiting after the burn had completed, using the differences between the input
and output states of the integration, rather than directly limiting the RHS in the ODE. The behavior of the
burn was qualitatively different, in that nickel production began much earlier in the collision when the limiting
was applied this way. Nevertheless, the qualitative final outcome of the collision -- only trace amounts
of nickel produced in total -- was still unchanged. With this variant too, it did not matter whether the limiting
factor was determined based on the initial value of the state or the final value of the state.
Our default resolution is significantly lower than that of \citeauthor{kushnir:2013} (the text
says they used resolutions up to a few km) but we have tried runs at higher resolution and
the results are qualitatively similar. We did not directly compare the results of the method at
the same resolution as theirs, but our results indicate at least that this limiter significantly
modifies the evolution for modest resolution simulations. The limiter does delay the time at which
the detonation occurs, as expected by \citeauthor{kushnir:2013}, but the conditions for generating
significant amounts of iron-group elements do not develop. It is not clear what the relevant difference
is between our simulation and that of \citeauthor{kushnir:2013}, but as the suppressed burn limiting
used here is ultimately unphysical, we choose not to use it in any other simulation.

@@

To answer these questions we performed a series of simulations at varying base
resolutions. In each case we ran the simulation to the point when the density
at the collision point reached $10^7\ \text{g / cm}^{-3}$, which we will informally
call the collision time, $t_{\rm c}$. The collision time is about $t_{\rm c} = 6.85\ \text{s}$
at high resolution. At low resolution, the simulation is slightly different because
of the distortion imposed by a rectangular grid representing initially spherical stars,
so the collision time is slightly later. Then, we restarted the run from this checkpoint
and reset the time to zero (such that $t_{\rm c} = 0$). The new simulation runs until about
$t = 2.15\ \text{s}$, so that the full 9 seconds are simulated. During this second phase
the detonation occurs. An estimate of the detonation time, $t_{\rm d}$, that serves well
is the time when the peak density on the grid first reaches $5 \times 10^9\ \text{K}$.
We measure the detonation time relative to $t_{\rm c}$. Approximately speaking, the later
$t_{\rm d}$ is, the more nickel is produced. As time passes, more white dwarf material
slams into the center, compressing and heating up the gas. Both effects are favorable
to production of nickel. So, a later $t_{\rm d}$ implies that more material can be converted
to nickel. Similarly, the speed of the detonation matters. A slower speed detonation prolongs
the collision event, allowing more gas to be processed into nickel.

>>>>

We define several burning modes that determine how $T$ and $e$ are evolved
during a nuclear burn. In a hydrostatic burn, which we call burning mode 0,
we keep $\rho$ and $T$ fixed throughout, and use 
the energy released at the end to compute a final temperature that is
thermodynamically consistent with the new internal energy. By contrast,
in a self-heating burn (mode 1), we allow the temperature to evolve in response
to the burning (see\footnote{In the cited paper, a term based on the
thermodynamic chemical potential was included; we now believe
that it is incorrect to include such a term in the burn, since it
automatically sums to zero analytically.} \citet{maestro3}):
\begin{equation}
  \frac{dT}{dt} = \frac{1}{c_V}\frac{\partial e}{\partial t}
\end{equation}
(Although $T$ evolves during the burn so that the integration is physically
accurate, as in the hydrostatic method we discard the final value
for $T$ at the end of the burn and recompute a temperature for the zone that is
consistent with its new internal energy.) Here $c_V$ is the specific heat at
constant volume, which is provided by the equation of state.  During this burn,
we can keep $c_V$ constant using its initial value, or at each step we
can choose to re-evaluate the equation of state using the latest value of $(\rho, T)$.
The latter is more expensive but also more accurate, and we use it in this paper.
In practice we find that the cost is small in comparison to the more expensive
parts of the calculation, and it can significantly speed up convergence near NSE.
A third option (mode 2) presented by \citet{raskin:2010} is a so-called ``hybrid'' mode.
In this mode, by default we do a hydrostatic burn. If that burn fails, or if the net
energy change is negative, we do the burn again in self-heating mode. A final option (mode 3)
is a burn that limits the changes due to a burn to avoid numerically unstable burning.
This mode is discussed in \autoref{sec:unstable_burning} and we will call it a
``suppressed'' burn for the remainder of this work. All four options
are implemented in our burner software.

Except for the discussion later about the suppressed burn mode, the simulations
shown in this work all use the self-heating mode. The hydrostatic and hybrid burns
are only a slight variation on the self-heating burn in practice for this problem.
The reason that the hydrostatic burn cannot substantially change the outcome is
that even though the temperature is fixed during the burn, it is \textit{not}
constant in the rest of the hydrodynamics step. And in the hydrodynamics step,
the \castro\ algorithm performs multiple EOS calls to ensure that the temperature
is synchronized with the internal energy at various points in the update. So in
practice there is a temperature change due to the burn for our problem. This seems
inevitable for any hydro code unless the hydrodynamics update includes an explicit
equation for $T$ that is independent of what is happening for the internal energy.
The hydrostatic burn does slightly limit the energy release during any given burn,
but summed up over the whole simulation this is not enough to qualitatively affect
the outcome. The hybrid burn does not affect the simulation outcome at all relative
to the hydrostatic burn. The difference between them only applies for zones that have
a net negative energy release from the hydrostatic burn, which occurs only after
we have burned to NSE. The hybrid burn was mainly implemented by \citet{raskin:2010}
for making it easier to numerically complete a burn, rather than because of the
effects it has on the outcome of the simulation.

@@

Our primary metric for this test is the amount of $^{56}$Ni generated in the collision,
as this is the parameter most directly related to the observable quantities of interest
for Type Ia supernovae. We collect information at the end of every timestep about the
total amount of nickel on the grid (in solar masses), and we use the maximum value of
this nickel mass. Plotfiles of relevant quantities are created every 0.01 seconds.
The timesteps are limited to ensure that every multiple of this plotfile interval is
reached exactly, which means an effective maximum timestep of 0.01 seconds.

@@

In \autoref{sec:burner} we observed that there is an alternative to the traditional
self-heating approach in a burn, the ``suppressed'' burning mode. This alternative
is relevant to a collision problem because, as pointed out by \cite{raskin:2010}
and \cite{kushnir:2013}, it is possible for a typical zone in our simulation to
release a very large amount of energy in a burn before it is cooled
off by a subsequent hydrodynamic expansion step, leading to the numerically unstable burning
problem described in \autoref{sec:unstable_burning}. However, the suppressed burning mode
(modeled after \citet{kushnir:2013}) produces only about
$2 \times 10^{-4}\ \msolar$ of $^{56}$Ni. In a suppressed burn, we ensure that at every evaluation
of the right-hand-side vector for the nuclear network integration, all quantities are scaled by the
same factor, with this factor ensuring that the energy release is not large enough to permit a
spurious detonation. This seems to be equivalent to the method used in \citet{kushnir:2013}. We
reproduce their claim that this suppresses the detonation until later in the collision, after more
material has approached the stalled shock. We also observe that, for a given resolution, the simultaneous
detonations occur slightly outside the initial contact point, near the edges of the stalled shock. They
establish as the main source of error in simulations that produce too little nickel. However, we find
that the resulting detonations are not strong enough to convert a significant amount of material to NSE
conditions; instead, only QSE conditions are reached near the center of the collision, leaving a significant
amount of silicon and sulfur on the domain that is not further processed. In order to ensure that the
very different results we obtained were not a result of the specific choice we made for how to suppress
the burn, we tried varying the approach discussed in \autoref{sec:burner}. First, we limited based on
the current value of the internal energy and sound speed rather than the fixed, initial value during a burn
(this should be a less severe limiter). This did not yield a significant difference in the outcome. Second,
we also tried applying the limiting after the burn had completed, using the differences between the input
and output states of the integration, rather than directly limiting the RHS in the ODE. The behavior of the
burn was qualitatively different, in that nickel production began much earlier in the collision when the limiting
was applied this way. Nevertheless, the qualitative final outcome of the collision -- only trace amounts
of nickel produced in total -- was still unchanged. With this variant too, it did not matter whether the limiting
factor was determined based on the initial value of the state or the final value of the state.
Our default resolution is significantly lower than that of \citeauthor{kushnir:2013} (the text
says they used resolutions up to a few km) but we have tried runs at higher resolution and
the results are qualitatively similar. We did not directly compare the results of the method at
the same resolution as theirs, but our results indicate at least that this limiter significantly
modifies the evolution for modest resolution simulations. The limiter does delay the time at which
the detonation occurs, as expected by \citeauthor{kushnir:2013}, but the conditions for generating
significant amounts of iron-group elements do not develop. It is not clear what the relevant difference
is between our simulation and that of \citeauthor{kushnir:2013}, but as the suppressed burn limiting
used here is ultimately unphysical, we choose not to use it in any other simulation.

@@

To answer these questions we performed a series of simulations at varying base
resolutions, augmented with additional runs with AMR. In each case we ran the
simulation to the point at which unstable burning was approached. Then we stopped
the run and created a checkpoint. Starting from that checkpoint, we did a simulation
at varying levels of refinement based on the unstable burning criterion. We will refer
to the refinement through the refinement factor, $r$. This factor is the jump in
resolution between the coarse grid and the finest grid. The uniform base grid case is $r = 1$.
Meanwhile, $r = 16$ means a uniform coarse grid plus two levels of 4x refinement.

The collision time is about $t_{\rm c} = 6.85\ \text{s}$
at high resolution. At low resolution, the simulation is slightly different because
of the distortion imposed by a rectangular grid representing initially spherical stars,
so the collision time is slightly later. Then, we restarted the run from this checkpoint
and reset the time to zero (such that $t_{\rm c} = 0$). The new simulation runs until about
$t = 2.15\ \text{s}$, so that the full 9 seconds are simulated. During this second phase
the detonation occurs. An estimate of the detonation time, $t_{\rm d}$, that serves well
is the time when the peak density on the grid first reaches $5 \times 10^9\ \text{K}$.
We measure the detonation time relative to $t_{\rm c}$. Approximately speaking, the later
$t_{\rm d}$ is, the more nickel is produced. As time passes, more white dwarf material
slams into the center, compressing and heating up the gas. Both effects are favorable
to production of nickel. So, a later $t_{\rm d}$ implies that more material can be converted
to nickel. Similarly, the speed of the detonation matters. A slower speed detonation prolongs
the collision event, allowing more gas to be processed into nickel.
